Do I Need to Model You in order to Learn With You?

### 2015-05-10 mutual-modeling.tex
Add missing '0' in stats values + a bit of house keeping

From its very beginning, CSCL research has been following
\citet{roschelle1995construction} suggestion that collaborative learning has
something to do with the process of constructing and maintaining a \emph{shared
understanding} of the task at hand. Building a shared/mutual understanding
refers to the upper class of collaborative learning situations, those in which
students should build upon each other's understanding to refine their own
understanding: what is expected to produce learning is not the mere fact that
two students build the same understanding but the cognitive effort they have to
engage to build this shared understanding~\citep{schwartz1995emergence}. This
effort can be observed by the frequency of rich interactions, \ie interactions
whose occurrence has been related to learning: (self-) explanations in cognitive
science~\citep{chi1989self, webb1991task}, conflict resolution in
socio-cognitive theories~\citep{doise1975social},
mutual regulation~\citep{blaye1995collaborative} in a Vygostkian perspective, etc. The
construction of a shared understanding has been investigated for several years
in psycholinguistics, under the  notion of
\emph{grounding}~\citep{clark1986referring}. However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in learning
sciences. The monitoring and repair of misunderstanding explains for instance
referential failures in short dialogue episodes but does hardly predict
\emph{conceptual change} (\ie the acquisition, acceptation and integration of a
new belief into one's mental model\fxnote{check with Pierre}) over longer
sessions~\cite{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

\paragraph{Grounding criterion} The grounding criterion was high: the
correlation between \gModel{A}{B} and task performance was 0.42, $p = 0.05$.
Pairs with an accurate mutual model reached higher scores. A regression analysis
confirmed the positive and significant relation between group performance and
mutual modelling accuracy ($\beta=54, p = .02$).

@@

\paragraph{Symmetry question} We computed intra-class correlation as described
by~\citet{kenny1998data} from the answers to the cross-questionnaires.
Considering all pairs in both conditions, we found a positive and significant
correlation ($r = .38, p < .05$) between \gModel{A}{B} and \gModel{B}{A}.
Interestingly, this was higher in the control group ($r = 0.44$) than in the
experimental group ($r = 0.24$). Actually, $\Delta(\gMdeg{A}{B},\gMdeg{B}{A})$,
\ie the absolute differences between the models accuracy, was not significantly
different with or without the awareness tool ($F [1,13]= 0.144, p > 0.5$). This
result could be explained by the fact that the players without awareness tools
communicated more.

@@

\paragraph{Symmetry question} The correlation between \gModel{A}{B}  and
\gModel{B}{A}  is positive ($r = .41$) and significant ($p < .01$): the more $A$
makes errors about $B$, the more $B$ does as well.

@@

\paragraph{Study-specific questions} Our hypothesis about media richness is
rejected: the average \Model{A}{B}{emotions} was .62 ($SD = .28$) with the
audio+video condition and 0.71 ($SD = 0.19$) in the audio condition ($F=3.4, p:
0.065$).

@@

\paragraph{Grounding criterion} In this task, the grounding criterion was low.
We did not however evaluate task performance (\eg the quality of the jointly
produced concept map) but learning gains. The correlation between \gModel{A}{B}
and $A$'s learning gains is not significant ($\beta =.08, ns, N = 60$). It is
also not significant within each condition.

\paragraph{Study-specific questions} We performed a non-parametric Mann-Whitney
test on post-test scores for electric questions and a one-way ANOVA on scores
for ionic questions (Levene tests for homogeneity of variances): $p = .02$ and
$ns$, respectively. Results  did not show any significant difference between the
\emph{same information} (SI) condition and the \emph{complementary information}
(CI) condition, neither for electric questions [$U = 388.50, z = -0.88, ns$] nor
for ionic questions [$F(1, 58) = 0.17, ns$].  The effect of scripts on
\gModel{A}{B} was not significant ($F(1, 58) = 0.78, ns$) when considering the
absolute difference between \gModel{A}{B} and $B$'s post-test score. However, A
tended underestimate $B$'s score in the SI condition ($M = -2.06$) and to
overestimate it in the CI condition ($M = 1.21$) ($F(1, 58) = 6.44, p<.01$).
Regarding to \gModel{A}{A}, there was no significant difference between
conditions.

\paragraph{Symmetry question} The inter-class correlation between \gModel{A}{B}
and \gModel{B}{A} is approaching significance ($r=.26, F(1,29)=1.71, p=
0.075$). It is indeed significant when students read the same text (SI
condition: $r=0.43, F(1,15)=2.53, p<.05$) but not when they read different texts
($r=.13, F(1,12)=1.3, ns$).

@@

In this experiment, the symmetry of mutual modelling is only found when subjects
are in conditions of symmetry of information. This consolidates the hypothesis
of \gModel{A}{B} being a property of interactions in context rather than $A$'s
cognitive attitude\fxfatal{to be better justified}.

@@

\paragraph{Grounding criterion} A linear regression revealed a positive relation
between \gModel{A}{B} and the learning gain of the pair $\{A, B\}$ (calculated
as the average of the individual learning gains): $\beta= .401, p < .001, r2adj.
=.15$, large effect. This relationship was not significant in the previous
experiment which as been conducted with the same task (using the same task
should produce the same grounding criterion). This is probably due to the fact that
\gModel{A}{B} was influenced by the KAT.

\paragraph{Study-specific questions} The $t$-test reported a significant
difference between the KAT condition participants ($M = 13.4\%$) and the control
group [$M = 3.6\%, t(1, 60) = 2.73, p < .01$, Cohen's $d = 0.7$, medium to large
effect]: providing learners with cues about the prior-knowledge of their partner
enhances their collaborative learning. As a treatment check, we found a positive
and significant correlation between the amount of gazes on KAT (using eye
tracking devices) and the learning gains ($r(22) = .54, p = .01$). A detailed
analysis revealed that subjects look at KAT to assess their peer's credibility
when he/she provided new information.

The KAT has a significant effect on \gModel{A}{B}: peers more accurately
estimated their partners knowledge ($M = 1.11$) than those in the control
condition [$M = .98, t(1, 60) = 3.19, p < .01$, Cohen's $d = 0.83$, large
effect]. This is a trivial result since the KAT provided them with an initial
\gmodel{A}{B}. However, subjects have to predict the post-test score while the
KAT informed them about the pre-test score. Actually, pairs in the KAT condition
produced significantly more instances of 3 specific categories of interactions:
(1) elaborated utterances with rich contents,  (2) utterances asking about the
other's knowledge such as ``Did you understand how transmission works?'' (3)
utterances describing one's own knowledge (\gmodel{A}{A}) such as ``I don't
remember the Ranvier's thing...''. These 3 categories provide different account
of \gModel{A}{B}: (1) as an effect of the quality of interaction, (2) as an
effect of $A$'s effort to model $B$ or (3) as $B$'s effort to give cues to $A$ about his
own knowledge.

We examined \gModel{A}{B} as potentially mediating the effect of the KAT factor
on the relative learning gain. A linear regression confirmed that \gModel{A}{B}
was significantly related to the KAT-factor ($\beta= .381, p < .01, r2 = .15$).
The KAT-factor was also significantly and positively related to the RLG ($\beta=
.332, p < .01, r2 = .11$). We then tested the relation between the independent
variable (KAT) and the dependent variable (gains) when controlling for the
mediating variable (\gModel{A}{B}). A multiple regression showed that the
KAT-factor was no longer a significant predictor ($\beta= .210; p = ns$) whereas
the \gModel{A}{B} was still a significant predictor ($\beta= .32, p < .01$).
Thus, it can be concluded that \gModel{A}{B} mediated the KAT-factor's effect on
the learners' RLG. The Sobel significance test for indirect effects was
significant [$z = 1.99, p < .05$].

@@

\begin{table*}[h!t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{p{3cm}|p{4cm}|p{4cm}|p{4cm}|p{4cm}|p{4cm}}
\textit{}                     & \textbf{Study 1}                 & \textbf{Study 2}                & \textbf{Study 3}          & \textbf{Study 4}          & \textbf{Study 5}             \\ \hline
\textit{Task}                 & Game in virtual space            & Game in physical space          & Building an argument map  & Building a concept map    & Building a concept map       \\
\textit{Interactions}         & Audio                            & Written                         & Audio / Video             & Audio                     & Audio                        \\
\textit{Shared editor}        & 3D Space                         & 2D Map                          & Concept Map               & Concept Map               & Concept Map                  \\
\textit{Group Size}           & 2                                & 3                               & 3                         & 3                         & 3                            \\
\textit{Duration (mean)}      & 90min                            & 16min                           & 61min                     & 90min                     & 90min                        \\
\textit{Awareness tool}       & Partner's position / orientation & Partner's current/past position & -                         & Partner's concept map     & Partner's scores at pre-test \\
\textit{Dependent var.}   & Team/Ind. performance            & Team/Ind. performance           & Team performance          & Team/Ind. knowledge gains & Team/Ind. knowledge gains    \\
    \textit{Independent var.} & Awareness tool \textit{vs} not.           & Awareness tool \textit{vs}  not.          & Audio \textit{vs} Audio+Video & Script \textit{vs} not.            & Awareness tool \textit{vs} not.       \\
\textit{MM°-DV correlation}   & +0.42                            & +0.15                           & +0.22                     & +0.08                     & +0.40                        \\
\textit{$\Delta_1$}           & +0.38 (significant)              & +0.41 (significant)             & ???                       & +0.26 (near significant)          & +0.05 (ns)              \\
\textit{$\Delta_2$}           &                                  & +0.43 (significant)             & +0.43                     &                           &                              \\
    \textit{$\Delta_3$}       &                                  & +0.30 (significant)             & +0.66                     &                           &                              \\
\textit{$\Delta_4$}           &                                  &                                 &                           & +0.05                     &                             
\end{tabular}
}
\caption{Comparison of the studies}
\label{synthesis_table}
\end{table*}

@@

Among these complex results, it is important to stress that, for $\Delta_2$, the
significant correlations are all below .50, not around .90.  So even if all
$\Delta_2$ correlations were significant in all experiments, we would still not
conclude that modelling purely reflects the quality of social interactions.
There is obviously a large part of individual variance; some of us are much
better in modelling peers. This is not new. The contribution of these studies is
rather to establish the importance of team processes on something that a priori
seemed to be more of a personal cognitive skill.  In other words, we stress that
the bottle is half-full.

>>>>

From its very beginning, CSCL research has been following
\citet{roschelle1995construction} suggestion that collaborative learning has
something to do with the process of constructing and maintaining a \emph{shared
understanding} of the task at hand. Building a shared/mutual understanding
refers to the upper class of collaborative learning situations, those in which
students should build upon each other's understanding to refine their own
understanding: what is expected to produce learning is not the mere fact that
two students build the same understanding but the cognitive effort they have to
engage to build this shared understanding~\citep{schwartz1995emergence}. This
effort can be observed by the frequency of rich interactions, \ie interactions
whose occurrence has been related to learning: (self-) explanations in cognitive
science~\citep{chi1989self, webb1991task}, conflict resolution in
socio-cognitive theories~\citep{doise1975social},
mutual regulation~\citep{blaye1995collaborative} in a Vygostkian perspective, etc. The
construction of a shared understanding has been investigated for several years
in psycholinguistics, under the  notion of
\emph{grounding}~\citep{clark1986referring}. However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in learning
sciences. The monitoring and repair of misunderstanding explains for instance
referential failures in short dialogue episodes but does hardly predict
\emph{conceptual change} (\ie the acquisition, acceptation and integration of a
new belief into one's mental model) over longer
sessions~\cite{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

\paragraph{Grounding criterion} The grounding criterion was high: the
correlation between \gModel{A}{B} and task performance was 0.42, $p = 0.05$.
Pairs with an accurate mutual model reached higher scores. A regression analysis
confirmed the positive and significant relation between group performance and
mutual modelling accuracy ($\beta=54, p = 0.02$).

@@

\paragraph{Symmetry question} We computed intra-class correlation as described
by~\citet{kenny1998data} from the answers to the cross-questionnaires.
Considering all pairs in both conditions, we found a positive and significant
correlation ($r = 0.38, p < 0.05$) between \gModel{A}{B} and \gModel{B}{A}.
Interestingly, this was higher in the control group ($r = 0.44$) than in the
experimental group ($r = 0.24$). Actually, $\Delta(\gMdeg{A}{B},\gMdeg{B}{A})$,
\ie the absolute differences between the models accuracy, was not significantly
different with or without the awareness tool ($F [1,13]= 0.144, p > 0.5$). This
result could be explained by the fact that the players without awareness tools
communicated more.

@@

\paragraph{Symmetry question} The correlation between \gModel{A}{B}  and
\gModel{B}{A}  is positive ($r = 0.41$) and significant ($p < 0.01$): the more $A$
makes errors about $B$, the more $B$ does as well.

@@

\paragraph{Study-specific questions} Our hypothesis about media richness is
rejected: the average \Model{A}{B}{emotions} was 0.62 ($SD = 0.28$) with the
audio+video condition and 0.71 ($SD = 0.19$) in the audio condition ($F=3.4, p:
0.065$).

@@

\paragraph{Grounding criterion} In this task, the grounding criterion was low.
We did not however evaluate task performance (\eg the quality of the jointly
produced concept map) but learning gains. The correlation between \gModel{A}{B}
and $A$'s learning gains is not significant ($\beta = 0.08, ns, N = 60$). It is
also not significant within each condition.

\paragraph{Study-specific questions} We performed a non-parametric Mann-Whitney
test on post-test scores for electric questions and a one-way ANOVA on scores
for ionic questions (Levene tests for homogeneity of variances): $p = 0.02$ and
$ns$, respectively. Results  did not show any significant difference between the
\emph{same information} (SI) condition and the \emph{complementary information}
(CI) condition, neither for electric questions ($U = 388.50, z = -0.88, ns$) nor
for ionic questions ($F(1, 58) = 0.17, ns$).  The effect of scripts on
\gModel{A}{B} was not significant ($F(1, 58) = 0.78, ns$) when considering the
absolute difference between \gModel{A}{B} and $B$'s post-test score. However, $A$
tended underestimate $B$'s score in the SI condition ($M = -2.06$) and to
overestimate it in the CI condition ($M = 1.21$) ($F(1, 58) = 6.44, p<0.01$).
Regarding to \gModel{A}{A}, there was no significant difference between
conditions.

\paragraph{Symmetry question} The inter-class correlation between \gModel{A}{B}
and \gModel{B}{A} is approaching significance ($r= 0.26, F(1,29)=1.71, p=
0.075$). It is indeed significant when students read the same text (SI
condition: $r=0.43, F(1,15)=2.53, p<0.05$) but not when they read different texts
($r= 0.13, F(1,12)=1.3, ns$).

@@

In this experiment, the symmetry of mutual modelling is only found when subjects
are in conditions of symmetry of information. This consolidates the hypothesis
of \gModel{A}{B} being a property of interactions in context rather than $A$'s
cognitive attitude.

@@

\paragraph{Grounding criterion} A linear regression revealed a positive relation
between \gModel{A}{B} and the learning gain of the pair $\{A, B\}$ (calculated
as the average of the individual learning gains): $\beta= 0.401, p < 0.001, r2adj.
= 0.15$, large effect. This relationship was not significant in the previous
experiment which as been conducted with the same task (using the same task
should produce the same grounding criterion). This is probably due to the fact that
\gModel{A}{B} was influenced by the KAT.

\paragraph{Study-specific questions} The $t$-test reported a significant
difference between the KAT condition participants ($M = 13.4\%$) and the control
group [$M = 3.6\%, t(1, 60) = 2.73, p < 0.01$, Cohen's $d = 0.7$, medium to large
effect]: providing learners with cues about the prior-knowledge of their partner
enhances their collaborative learning. As a treatment check, we found a positive
and significant correlation between the amount of gazes on KAT (using eye
tracking devices) and the learning gains ($r(22) = 0.54, p = 0.01$). A detailed
analysis revealed that subjects look at KAT to assess their peer's credibility
when he/she provided new information.

The KAT has a significant effect on \gModel{A}{B}: peers more accurately
estimated their partners knowledge ($M = 1.11$) than those in the control
condition [$M = 0.98, t(1, 60) = 3.19, p < 0.01$, Cohen's $d = 0.83$, large
effect]. This is a trivial result since the KAT provided them with an initial
\gmodel{A}{B}. However, subjects have to predict the post-test score while the
KAT informed them about the pre-test score. Actually, pairs in the KAT condition
produced significantly more instances of 3 specific categories of interactions:
(1) elaborated utterances with rich contents,  (2) utterances asking about the
other's knowledge such as ``Did you understand how transmission works?'' (3)
utterances describing one's own knowledge (\gmodel{A}{A}) such as ``I don't
remember the Ranvier's thing...''. These 3 categories provide different account
of \gModel{A}{B}: (1) as an effect of the quality of interaction, (2) as an
effect of $A$'s effort to model $B$ or (3) as $B$'s effort to give cues to $A$ about his
own knowledge.

We examined \gModel{A}{B} as potentially mediating the effect of the KAT factor
on the relative learning gain. A linear regression confirmed that \gModel{A}{B}
was significantly related to the KAT-factor ($\beta= 0.381, p < 0.01, r2 = 0.15$).
The KAT-factor was also significantly and positively related to the RLG ($\beta=
.332, p < 0.01, r2 = 0.11$). We then tested the relation between the independent
variable (KAT) and the dependent variable (gains) when controlling for the
mediating variable (\gModel{A}{B}). A multiple regression showed that the
KAT-factor was no longer a significant predictor ($\beta= 0.210; p = ns$) whereas
the \gModel{A}{B} was still a significant predictor ($\beta= 0.32, p < 0.01$).
Thus, it can be concluded that \gModel{A}{B} mediated the KAT-factor's effect on
the learners' RLG. The Sobel significance test for indirect effects was
significant [$z = 1.99, p < 0.05$].

@@

\begin{table*}[h!t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{p{3cm}|p{4cm}|p{4cm}|p{4cm}|p{4cm}|p{4cm}}
\textit{}                     & \textbf{Study 1}                 & \textbf{Study 2}                & \textbf{Study 3}          & \textbf{Study 4}          & \textbf{Study 5}             \\ \hline
\textit{Task}                 & Game in virtual space            & Game in physical space          & Building an argument map  & Building a concept map    & Building a concept map       \\
\textit{Interactions}         & Audio                            & Written                         & Audio / Video             & Audio                     & Audio                        \\
\textit{Shared editor}        & 3D Space                         & 2D Map                          & Concept Map               & Concept Map               & Concept Map                  \\
\textit{Group Size}           & 2                                & 3                               & 3                         & 3                         & 3                            \\
\textit{Duration (mean)}      & 90min                            & 16min                           & 61min                     & 90min                     & 90min                        \\
\textit{Awareness tool}       & Partner's position / orientation & Partner's current/past position & -                         & Partner's concept map     & Partner's scores at pre-test \\
\textit{Dependent var.}       & Team/Ind. performance            & Team/Ind. performance           & Team performance          & Team/Ind. knowledge gains & Team/Ind. knowledge gains    \\
\textit{Independent var.}     & Awareness tool \textit{vs} none  & Awareness tool \textit{vs} none & Audio \textit{vs} Audio+Video & Script \textit{vs} none & Awareness tool \textit{vs} none       \\
\textit{MM°-DV correlation}   & +0.42                            & +0.15                           & +0.22                     & +0.08                     & +0.40                        \\
\textit{$\Delta_1$}           & +0.38 (significant)              & +0.41 (significant)             & ???                       & +0.26 (near significant)          & +0.05 (ns)              \\
\textit{$\Delta_2$}           &                                  & +0.43 (significant)             & +0.43                     &                           &                              \\
    \textit{$\Delta_3$}       &                                  & +0.30 (significant)             & +0.66                     &                           &                              \\
\textit{$\Delta_4$}           &                                  &                                 &                           & +0.05                     &                             
\end{tabular}
}
\caption{Comparison of the studies}
\label{synthesis_table}
\end{table*}

@@

Among these complex results, it is important to stress that, for $\Delta_2$, the
significant correlations are all below 0.50, not around 0.90.  So even if all
$\Delta_2$ correlations were significant in all experiments, we would still not
conclude that modelling purely reflects the quality of social interactions.
There is obviously a large part of individual variance; some of us are much
better in modelling peers. This is not new. The contribution of these studies is
rather to establish the importance of team processes on something that a priori
seemed to be more of a personal cognitive skill.  In other words, we stress that
the bottle is half-full.
